---
title: "healthcare_costs_kaggle_comp"
author: "Graham Chalfant"
date: "8/1/2021"
output: html_document
---

```{r setup, include=FALSE}

library(tidyverse)
library(stringr)
library(lubridate)
library(FSelector)

```

```{r}
#https://www.kaggle.com/ravichaubey1506/healthcare-cost#

data <- read_csv("HospitalCosts.csv")

#Check data is in correct form 
str(data)

#FEMALE needs to be categorical 
data$FEMALE <- as.factor(data$FEMALE)
data$RACE <- as.factor(data$RACE)
#data$APRDRG <- as.factor(data$APRDRG)

#Check NAs
summary(data)#There is one NA in RACE - will remove column 
count(data)

#Remove NAs
data <- na.omit(data) #removed the row with a NA race
count(data)

#Create age groups 
data <- data %>% mutate(AGE_GROUP = ifelse(between(data$AGE, 0, 5), "0-5", #Checks if the first expression is true. If so, then "0-5" is printed, if not, the next ifelse is tested
                                      ifelse(between(data$AGE, 6, 10), "6-10",
                                             ifelse(between(data$AGE, 11, 15), "11-15", "16-20"))))
#Factor the age groups
data$AGE_GROUP <- as.factor(data$AGE_GROUP)

#Assign levels to the new age categories 
levels(data$AGE_GROUP) <- c("0-5", "6-10", "11-15", "16-20")


```


\#Goals of the analysis


Question 1: To record the patient statistics, the agency wants to find the age category
of people who frequent the hospital and has the maximum expenditure.

Question 2: In order of severity of the diagnosis and treatments and to find out the
expensive treatments, the agency wants to find the diagnosis related group
that has maximum hospitalization and expenditure.

Question 3: To make sure that there is no malpractice, the agency needs to analyze if
the race of the patient is related to the hospitalization costs.

Question 4: To properly utilize the costs, the agency has to analyze the severity of the
hospital costs by age and gender for proper allocation of resources.

Question 5: Since the length of stay is the crucial factor for inpatients, the agency wants
to find if the length of stay can be predicted from age, gender, and race.

Question 6: To perform a complete analysis, the agency wants to find the variable that
mainly affects the hospital costs.


| Column Name | Description                                             |
|-------------|---------------------------------------------------------|
| AGE         | Age of the patient discharged (0-17)                    |
| FEMALE      | Binary variable that indicates if the patient is female |
| LOS         | Length of stay, in days                                 |
| RACE        | Race of the patient (specified numerically)             |
| TOTCHG      | Hospital discharge costs                                |
| APRDRG      | All patent refined diagnosis related groups             |
|             |                                                         |

: Data Dictionary


Data Exploration 
```{r}
hist(data$AGE)#mostly younger individuals 

table(data$FEMALE) #Even split of male and female - more females than males 

boxplot(data$LOS)#Could be potential outliers in length of stay - will determine after modeling 

data <- data %>% filter(LOS <= 10)
count(data) #removed 10 rows 

table(data$RACE) #Race is predominantly 1

boxplot(data$TOTCHG)
data <- data %>% filter(TOTCHG <= 40000) #Removing outliers

data %>% group_by(APRDRG) %>% count(APRDRG) %>% arrange(desc(n)) #APRDRG 640 is by far the most frequent diagnosis group 

```


## Question 1: To record the patient statistics, the agency wants to find the age category of people who frequent the hospital and has the maximum expenditure.

Average length of stay by age group
- Average stay decreases with age_group
```{r}
#Average stay by age 
average_stay_data <- data %>% group_by(AGE_GROUP) %>% summarise(avg_stay = mean(LOS)) %>% arrange(desc(avg_stay)) 

#Filters data in decending order based on LOS
#average_stay_data$AGE <- factor(average_stay_data$AGE, levels = average_stay_data$AGE[order(average_stay_data$avg_stay)])

ggplot(average_stay_data, aes(x = AGE_GROUP, y = avg_stay)) + geom_bar(stat = "identity", fill = "steelblue") + labs(title = "Average stay by age", x = "Age", y = "Average length of stay") #+ scale_x_continuous(breaks=seq(0,17,1))

```

Average and max expenditure by age 
- Average expenditure increases with age
- Max expenditure is greatest for age groups 6 - 15
```{r}
#Max discharge and average discharge by age
max_exp <- data %>% group_by(AGE_GROUP) %>% summarise(max_exp = max(TOTCHG), avg_exp = mean(TOTCHG)) %>% arrange(desc(max_exp))

max_exp_longer <- pivot_longer(max_exp, c(avg_exp, max_exp))

ggplot(max_exp_longer, aes(x = AGE_GROUP, y = value, fill = factor(name))) + geom_bar(stat = "identity", position = "dodge")+
  scale_fill_discrete(name="Expenditure",
                      labels=c("Average", "Max")) +
  labs(title = "Expenditure by age", x = "Age", y = "Expense")

```


## Question 2: In order of severity of the diagnosis and treatments and to find out the expensive treatments, the agency wants to find the diagnosis related group that has maximum hospitalization and expenditure.

Find the diagnosis related group that has maximum hospitalization and expenditure.
```{r}
max_diagnosis <- data %>% group_by(APRDRG) %>% summarise(MAX_CHARGE = max(TOTCHG)) %>% slice_max(MAX_CHARGE, n = 10)

ggplot(max_diagnosis, aes(x = reorder(APRDRG, -MAX_CHARGE), y = MAX_CHARGE, fill = MAX_CHARGE)) + geom_bar(stat = "identity") + labs(title = "Max Charge by Diagnosis", x = "Diagnosis", y = "Max Charge") + theme(legend.position = "none") + geom_text(aes(label = MAX_CHARGE), vjust=1.6, color="white", size=3.5)

```


## Question 3: To make sure that there is no malpractice, the agency needs to analyze if the race of the patient is related to the hospitalization costs.

Seems that there is potential for malpractice. Race two's average costs are almost double the average costs for all races. 
```{r}
race_plot <- data %>% group_by(RACE) %>% summarise(AVG_COST = mean(TOTCHG))

total_average <- mean(data$TOTCHG)

ggplot(race_plot, aes(x = RACE, y = AVG_COST)) + geom_bar(stat = "Identity", fill = "steelblue") + labs(title = "Average Cost by Race", y = "Average Cost", x = "Race") + geom_hline(aes(yintercept = total_average, color = "red")) + scale_fill_discrete(name = "Average", labels = 2440) + theme(legend.position = "none")

```


```{r}
summary(lm(TOTCHG~RACE, data)) # Race is not a predictor of costs
```


## Question 4: To properly utilize the costs, the agency has to analyze the severity of the hospital costs by age and gender for proper allocation of resources.
- Average costs increase for males with age
- Average costs are much lower for females, and slightly increase with age until 16-20, where it reaches its lowest point

```{r}
age_gender_costs <- data %>% group_by(AGE_GROUP, FEMALE) %>% summarise(avg_costs = mean(TOTCHG)) %>% arrange(desc(avg_costs))

ggplot(age_gender_costs, aes(x = AGE_GROUP, y = avg_costs, fill = factor(FEMALE))) + geom_bar(stat = "Identity", position = "dodge") + scale_fill_discrete(name = "Gender", labels = c("Male", "Female")) + labs(title = "Average Costs by Gender and Age Group", x = "Age Group", y = "Average Costs")

```

```{r}
summary(lm(TOTCHG~AGE + FEMALE, data)) # AGE and FEMALE have a significant impact on TOTCHG based on p-values
```


## Question 5: Since the length of stay is the crucial factor for inpatients, the agency wants to find if the length of stay can be predicted from age, gender, and race.

Age, gender, and race are poor predictors of LOS
```{r}
lm1 <- lm(LOS ~ AGE + FEMALE + RACE, data)

summary(lm1)

```

```{r}
#Calculating information gain 
attribute_weights <- information.gain(LOS ~., data)

#Ordering attributes based on IG
attribute_weights %>% arrange(desc(attr_importance))
```

New model based on IG
- TOTCHG, APRDRG, and AGE are all significant predictors of LOS
```{r}
data$APRDRG <- as.numeric(data$APRDRG) #Change back to numeric for model

lm2 <- lm(LOS ~ TOTCHG + APRDRG + AGE, data)

summary(lm2)
```


## Question 6: To perform a complete analysis, the agency wants to find the variable that mainly affects the hospital costs.

```{r}
#Calculating information gain 
attribute_weights2 <- information.gain(TOTCHG ~., data)

#Ordering attributes based on IG
attribute_weights2 %>% arrange(desc(attr_importance))
```

- LOS and APRDRG are significant based on p-values
- Adjusted R-squared: 53.57 percent
- Cost is positively correlated with LOS 
```{r}
lm3 <- lm(TOTCHG~LOS+APRDRG, data) #Regression with LOS and APRDRG as predictor variables 

summary(lm3)
```





